---
title: "Untitled"
author: "Fariba"
date: "2023-05-15"
output:
  html_document:
    df_print: paged
---


```{r}
getwd()
setwd("E:/Bolton university/data science/assignment2")
data <- read.csv("data.csv", header = FALSE)
data
```

## Reshape the data
Before going further, we reshape data to better work with the data.
In order to do this, first, we keep the first two columns which are the locations:
```{r}
# Keep the first and second columns in variable location
location <- data[, 1:2]
location
```


And then we remove these columns from our dataset:
```{r}
# Remove the first and second columns from the data frame
data <- subset(data, select = -c(1, 2))
# Remove the row names from the data frame
rownames(data) <- NULL
data
```

Now, we reshape the data and add a new column for time:
```{r reshape}
# create an empty list to store the reshaped data frames
data_list <- list()

# loop over each row of the original time series
for (i in 1:nrow(data)) {
  
  # extract the current row and convert to matrix
  current_row <- as.matrix(data[i,])
  
  # reshape the current row and add column names
  current_reshaped <- matrix(as.numeric(current_row), nrow = 248, ncol = 10, byrow = TRUE)
  colnames(current_reshaped) <- c("TSK","PSFC","U10","V10","Q2","Rainc","Rainnc","Snow","TSLB","SMOIS")

  # add a date column to the reshaped data
  current_df <- cbind.data.frame(date = seq(as.POSIXct("2018-05-01 00:00:00"), as.POSIXct("2018-05-31 21:00:00"),
                                            by = "3 hours"), current_reshaped)
  
  # add the current data frame to the list
  data_list[[i]] <- current_df
}

# combine all data frames in the list into a single data frame
final_df <- do.call(rbind, data_list)
final_df
```

After that, we add location columns to our dataset
```{r}
# Repeat each location coordinate 248 times
locations_rep <- data.frame(rep(location[[1]], each = 248), rep(location[[2]], each = 248))


colnames(locations_rep) <- c("Latitude", "Longitude")

# Merge final_df and locations_rep data frames
df_with_locations <- cbind(final_df, locations_rep)

df_with_locations
```


And then we move the last two columns to the first:
```{r}
# moving the last two columns to the first
df <- df_with_locations[, c(ncol(df_with_locations)-1, ncol(df_with_locations), 1:(ncol(df_with_locations)-2))]
df
```

## EDA(Exploratory Data Analysis)
```{r}
str(df)
```
```{r}
summary(df)
```
```{r}
head(df)
```
```{r}
tail(df)
```

### Check for missing values
```{r}
sum(is.na(df))
```

### Visualizing the data
```{r warning=FALSE}
library(ggplot2)
ggplot(df , aes(x = U10)) + 
   geom_histogram(binwidth = 2, fill ="blue", alpha = 0.7) + 
   labs(title = "Histogram of U10", x = "X component of wind at 10m", y = "frequency")
```
```{r warning=FALSE}
library(ggplot2)
ggplot(df , aes(x = V10)) + 
   geom_histogram(binwidth = 2, fill ="blue", alpha = 0.7) + 
   labs(title = "Histogram of V10", x = "Y component of wind at 10m", y = "frequency")
```
```{r warning=FALSE}
ggplot(df, aes(x = U10, y = V10)) + 
   geom_point(color = "blue") + 
   labs(x = "U10", y = "V10")
```


## Data cleaning
### Check duplicates
```{r}
sum(duplicated(df))
```
```{r}
sum(is.na(df))
```
```{r include=FALSE}
library(imputeTS)
```

```{r warning=FALSE}
df <- na.interpolation(df)
df
```
```{r}
sum(is.na(df))
```

## Create a new column for wind speed
```{r}
df$wind_speed <- sqrt(df$U10^2 + df$V10^2)
df
```

## detecting outliers
```{r}
ggplot(df, aes(y = wind_speed))+
  geom_boxplot(fill= "blue", alpha = 0.7)+
  labs(title = "Box Plot of wind_speed", y = "wind_speed")
```
```{r include=FALSE}
library(dplyr)
```
```{r}
df %>%
  select(wind_speed) %>%
  summarise(
    lower_extremes = quantile(wind_speed, 0),
    lower_quartile = quantile(wind_speed, 0.25),
    median = quantile(wind_speed, 0.5),
    upper_quartile = quantile(wind_speed, 0.75),
    upper_extremes = quantile(wind_speed, 1),
    random_quartile = quantile(wind_speed, 0.8)
  )
```
```{r}
data <- df$wind_speed
z_scores <- scale(data)
threshold <- 3
z_outliers <- data[abs(z_scores) > threshold]
print(z_outliers)
```

## Replace the outliers with the Winsorized values
```{r include=FALSE}
library(DescTools)
```

```{r}
winsorized_data <- Winsorize(data, probs = c(0.05, 0.95))
data[z_outliers] <- winsorized_data[z_outliers]
data <- winsorized_data
head(data)
```
```{r}
boxplot(data, main = "Box Plot of wind_speed after Handling Outliers")
```
```{r}
df$wind_speed <- data
df
```

## Keep 2 columns
```{r}
df %>%
  filter(Latitude  == 56.781 & Longitude == 3.508) %>%
  select(date, wind_speed) -> df
df
```

```{r}
str(df)
```

## Time series plot
```{r}
ggplot(df, aes(x = date, y = wind_speed)) +
  geom_line(color = "blue") +
  labs(title = "wind speed Over Time",
       x = "Date and Time",
       y = "wind speed") +
  theme_minimal()
```
```{r}
df <- df %>%
  mutate(time = as.numeric(difftime(date, min(date), units = "hours")))
df
```

## Split the data into training and test sets
```{r}
set.seed(123)
train_indices <- sample(1:nrow(df), 0.8 * nrow(df))
train_data <- df[train_indices, ]
test_data <- df[-train_indices, ]
str(train_data)
str(test_data)
```

## Model1: ARIMA
```{r include=FALSE}
library(forecast)
```

```{r}
#fit the ARIMA model
arima_model <- auto.arima(train_data$wind_speed)
summary(arima_model)
```
```{r}
# forecast
arima_forecast <- forecast(arima_model, h = length(test_data$wind_speed))

# evaluate
arima_accuracy <- accuracy(arima_forecast, test_data$wind_speed)
cat("ARIMA Model Accuracy:", arima_accuracy[, 'RMSE'])
```

## Model2: Linear regression
```{r}
# Fit the model on the training set
train_model <- lm(wind_speed ~ time, data = train_data)
summary(train_model)

# Predict wind_speed values for the test set
predictions <- predict(train_model, newdata = test_data)

# Calculate the root mean squared error (RMSE)
rmse_lr <- sqrt(mean((test_data$wind_speed - predictions)^2))
cat("RMSE_lr:", rmse_lr)
```
```{r}
#Plot Actual vs Predicted values
ggplot() +
  geom_point(data = test_data, aes(x = wind_speed, y = predictions), color = "blue") +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Actual vs. Predicted wind_speed",
       x = "Actual wind_speed",
       y = "Predicted wind_speed") +
  theme_minimal()
```

## Model3: SVR
```{r warning=FALSE, include=FALSE}
library(e1071)
```

```{r warning=FALSE}
# Fit an SVR model on the training set
# 1.Radial basis function (RBF) kernel
svr_model_RBF <- svm(wind_speed ~ time, data = train_data, kernel = "radial")
# Display the SVR model summary
summary(svr_model_RBF)
```
```{r warning=FALSE}
# 2.Linear kernel
svr_model_linear <- svm(wind_speed ~ time, data = train_data, kernel = "linear")
# Display the SVR model summary
summary(svr_model_linear)
```
```{r warning=FALSE}
# 3.Polynomial kernel
svr_model_poly <- svm(wind_speed ~ time, data = train_data, kernel = "polynomial")
# Display the SVR model summary
summary(svr_model_poly)
```
```{r warning=FALSE}
# Predict wind_speed values for the test set using the SVR models
svr_predictions_RBF <- predict(svr_model_RBF, newdata = test_data)
svr_predictions_linear <- predict(svr_model_linear, newdata = test_data)
svr_predictions_poly <- predict(svr_model_poly, newdata = test_data)

# Calculate the root mean squared error (RMSE) for the SVR models
svr_rmse_RBF <- sqrt(mean((test_data$wind_speed - svr_predictions_RBF)^2))
svr_rmse_linear <- sqrt(mean((test_data$wind_speed - svr_predictions_linear)^2))
svr_rmse_poly <- sqrt(mean((test_data$wind_speed - svr_predictions_poly)^2))
cat("SVR_RBF RMSE:", svr_rmse_RBF, "\n")
cat("SVR_linear RMSE:", svr_rmse_linear, "\n")
cat("SVR_poly RMSE:", svr_rmse_poly, "\n")
```
```{r warning=FALSE, include=FALSE}
library(gridExtra)
```

```{r}

# Plot the actual vs. predicted values for the SVR models
p1 <- ggplot() +
  geom_point(data = test_data, aes(x = wind_speed, y = svr_predictions_RBF), color = "blue") +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "SVR_RBF: Actual vs. Predicted wind_speed",
       x = "Actual wind_speed",
       y = "Predicted wind_speed") +
  theme_minimal()+
  theme(text = element_text(size = 8))

p2 <- ggplot() +
  geom_point(data = test_data, aes(x = wind_speed, y = svr_predictions_linear), color = "blue") +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "SVR_linear: Actual vs. Predicted wind_speed",
       x = "Actual wind_speed",
       y = "Predicted wind_speed") +
  theme_minimal()+
  theme(text = element_text(size = 8))

p3 <- ggplot() +
  geom_point(data = test_data, aes(x = wind_speed, y = svr_predictions_poly), color = "blue") +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "SVR_poly: Actual vs. Predicted wind_speed",
       x = "Actual wind_speed",
       y = "Predicted wind_speed") +
  theme_minimal()+
  theme(text = element_text(size = 8))

grid.arrange(p1, p2, p3, ncol = 1)
```

## Model4: Random forest
```{r warning=FALSE, include=FALSE}
library(randomForest)
```
```{r}
# Fit a Random Forest model on the training set
rf_model_n100 <- randomForest(wind_speed ~ time, data = train_data, ntree = 100)
# Display the Random Forest model summary
summary(rf_model_n100)
```
```{r}
# Fit a Random Forest model on the training set
rf_model_n200 <- randomForest(wind_speed ~ time, data = train_data, ntree = 200)
# Display the Random Forest model summary
summary(rf_model_n200)
```
```{r}
# Fit a Random Forest model on the training set
rf_model_n300 <- randomForest(wind_speed ~ time, data = train_data, ntree = 300)
# Display the Random Forest model summary
summary(rf_model_n300)
```
```{r}
# Predict wind_speed values for the test set using the Random Forest model
rf_predictions_n100 <- predict(rf_model_n100, newdata = test_data)
rf_predictions_n200 <- predict(rf_model_n200, newdata = test_data)
rf_predictions_n300 <- predict(rf_model_n300, newdata = test_data)

# Calculate the root mean squared error (RMSE) for the Random Forest model
rf_rmse_n100 <- sqrt(mean((test_data$wind_speed - rf_predictions_n100)^2))
rf_rmse_n200 <- sqrt(mean((test_data$wind_speed - rf_predictions_n200)^2))
rf_rmse_n300 <- sqrt(mean((test_data$wind_speed - rf_predictions_n300)^2))
cat("Random Forest n100 RMSE:", rf_rmse_n100, "\n")
cat("Random Forest n200 RMSE:", rf_rmse_n200, "\n")
cat("Random Forest n300 RMSE:", rf_rmse_n300, "\n")
```
```{r}
# Plot the actual vs. predicted values for the Linear Regression, SVR, and Random Forest models
p1 <- ggplot() +
  geom_point(data = test_data, aes(x = wind_speed, y = rf_predictions_n100), color = "blue") +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Random Forest: Actual vs. Predicted wind_speed",
       x = "Actual wind_speed",
       y = "Predicted wind_speed") +
  theme_minimal()+
  theme(text = element_text(size = 8))

p2 <- ggplot() +
  geom_point(data = test_data, aes(x = wind_speed, y = rf_predictions_n200), color = "blue") +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Random Forest: Actual vs. Predicted wind_speed",
       x = "Actual wind_speed",
       y = "Predicted wind_speed") +
  theme_minimal()+
  theme(text = element_text(size = 8))

p3 <- ggplot() +
  geom_point(data = test_data, aes(x = wind_speed, y = rf_predictions_n300), color = "blue") +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Random Forest: Actual vs. Predicted wind_speed",
       x = "Actual wind_speed",
       y = "Predicted wind_speed") +
  theme_minimal()+
  theme(text = element_text(size = 8))

grid.arrange(p1, p2, p3, ncol = 1)
```
```{r}
rmse_values <- c(arima_accuracy[ ,'RMSE'], rmse_lr, svr_rmse_RBF,svr_rmse_linear,svr_rmse_poly,rf_rmse_n100,rf_rmse_n200,rf_rmse_n300)
names = c("ARIMA_trainig", "ARIMA_test", "Linear Regression", "SVR RBF", "SVR Linear", "SVR Poly", "RF 100", "RF 200", "RF 300")
barplot(rmse_values, names.arg = names, xlab = "Model", ylab = "RMSE",main = "RMSE for Different Models", legend = names, col = c("red", "blue", "green", "yellow", "purple", "orange", "pink", "gray"),ylim = c(0, 3))
```

